Spike-and-Slab LASSO Generalized Additive Models and Scalable Algorithms for High-Dimensional Data Analysis

### 2022-05-16 Manuscript/01-Intro.Rmd
remove revision highlights in red

Much modern biomedical research, e.g., sequencing data analysis and electronic health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see Mallick and Yi [@Mallick2013] for review. These models are built upon a restrictive and unrealistic assumption, linearity. In classical statistical modeling, many strategies and models are proposed to relax the linearity assumption with various degrees of complexity. For example, variable categorization is a simple and common practice in epidemiology but suffers from power and interpretation issues. More complex models to address nonlinear effects include random forest and other so-called "black box" models [@Breiman2001]. These models are useful for statistical prediction but do not estimate parameters relevant to the data generation process that one can draw inferences from. In addition, how to generalize these \bg{"black box"} models to the high-dimensional setting remains unclear.

\bg{For their straightforward interpretation and flexibility, nonparametric regression models serve as great alternatives to the "black-box" models in prediction and variable selection}. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, which is assumed to follow some exponential family distribution, can be modeled with the summation of smooth functions. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis.

@@

\bg{We propose a novel Bayesian hierarchical generalized additive model (BHAM) for outcome prediction in the context of high-dimensional data analysis. Specifically, we incorporate smoothing penalties, derived from the smoothing spline literature \cite{Wood2017}, via reparameterization of smooth functions to avoid excessive shrinkage on the bases. Smoothing penalties were also previously used in the spike-and-slab GAM \cite{Scheipl2012} and the sparsity-smoothness penalty \cite{Meier2009}. We then impose a new two-part spike-and-slab LASSO prior to address the signal sparsity. In addition, a scalable optimization-based algorithm, EM-Coordinate Descent (EM-CD) algorithm, is developed. While the primary focus of this model is to improve prediction, the proposed model also provides utility in functional selection. Notably, the two-part prior that follows the effect hierarchy principle motivates a bi-level selection, rendering one of three possibilities for each predictor: no effect, only linear effect, or linear and nonlinear effects. The proposed model is implemented in a publicly available R package \texttt{BHAM} via \url{https://github.com/boyiguo1/BHAM}.}

>>>>

Much modern biomedical research, e.g., sequencing data analysis and electronic health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see Mallick and Yi [@Mallick2013] for review. These models are built upon a restrictive and unrealistic assumption, linearity. In classical statistical modeling, many strategies and models are proposed to relax the linearity assumption with various degrees of complexity. For example, variable categorization is a simple and common practice in epidemiology but suffers from power and interpretation issues. More complex models to address nonlinear effects include random forest and other so-called "black box" models [@Breiman2001]. These models are useful for statistical prediction but do not estimate parameters relevant to the data generation process that one can draw inferences from. In addition, how to generalize these "black box" models to the high-dimensional setting remains unclear.

For their straightforward interpretation and flexibility, nonparametric regression models serve as great alternatives to the "black-box" models in prediction and variable selection. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, which is assumed to follow some exponential family distribution, can be modeled with the summation of smooth functions. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis.

@@

We propose a novel Bayesian hierarchical generalized additive model (BHAM) for outcome prediction in the context of high-dimensional data analysis. Specifically, we incorporate smoothing penalties, derived from the smoothing spline literature \cite{Wood2017}, via reparameterization of smooth functions to avoid excessive shrinkage on the bases. Smoothing penalties were also previously used in the spike-and-slab GAM \cite{Scheipl2012} and the sparsity-smoothness penalty \cite{Meier2009}. We then impose a new two-part spike-and-slab LASSO prior to address the signal sparsity. In addition, a scalable optimization-based algorithm, EM-Coordinate Descent (EM-CD) algorithm, is developed. While the primary focus of this model is to improve prediction, the proposed model also provides utility in functional selection. Notably, the two-part prior that follows the effect hierarchy principle motivates a bi-level selection, rendering one of three possibilities for each predictor: no effect, only linear effect, or linear and nonlinear effects. The proposed model is implemented in a publicly available R package \texttt{BHAM} via \url{https://github.com/boyiguo1/BHAM}.
